---
title: "Assignment #7 Write-Up - Sumer Vaid "
output: github_document
---
```{r, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(poliscidata)

data(gss, package = "poliscidata")

# convert to tibble
gss <- as_tibble(gss)
```


In this exploratory data analysis, I attempt to formulate several hypothesis and attempt to validate them using exploratory data techniques. In this write-up, I will discuss three streamsof my most interesting findings. 

###Income, Hard-Work and Race

I was interested by people's prioritization of hard-work and how this related with their income levels. First,  I explored the general patterns of the importance assigned to hard-work: 

```{r, echo=FALSE}
hard_work<-ggplot(aes(workhard), data = gss) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+labs(x="Importance of Hardwork", y="Count")+ggtitle("How Important Is Hard-work?")
print(hard_work)
ggsave("hard_work.jpg")
```

I then explored how people's assigned importance of hard-work varied with their income levels. I expected to find that people who prioritized hard-work more were also high earners. 

```{r, echo=FALSE}
hard_income<-ggplot(aes(workhard,income06), data = gss) + geom_count() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+ggtitle("Relationship Between Hard-work and Income")+labs(x="Importance of Hardwork", y="Income Levels")
print(hard_income)
ggsave("hard_income.jpg")
```

My hypothesis was somewhat correct: people who rate hard-work as being least important clearly make lesser money than those who rate hard-work as being more than least important. However, instead of seeing a robust characterization of increasing hardwork with increasing income levels, I instead noted a bulk in the upper-middle income levels for most categorizations of hardwork. This means that middle-to-upper middle class families mostly rate hard-work as being atleast somewhat important. However, I was curious to see how this relationship is influenced by race. I assumed that members of low racial status would rate hard-work as being more important to them. 

```{r, echo=FALSE}
hard_income_race<-ggplot(aes(workhard,income06), data = gss) + geom_jitter() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+ggtitle("Relationship Between Hard-work, Income and Race")+labs(x="Importance of Hardwork", y="Income Levels")+facet_wrap(~race)
print(hard_income_race)
ggsave("hard_income_race.jpg")
```

My hypothesis was wrong: it looks like the relationship between importance assigned to hardwork and income levels is more robust for high-racial status members (Whites) as compared to Blacks. Lastly, I was intrigued by the missingness in the data. I decided to conduct some analysis to understand exactly who opted out of the hard-work question on the survey. I first examined the income levels of non-responders to the hard-work question. My hypothesis was that mostly wealthy individuals, who had acquired their wealth through means other than hardwork (nepotism), would not answer this question. 

```{r, echo=FALSE}
missing_hardwork<- subset(gss,is.na(workhard))
missingh_incomes<- ggplot(aes(income06), data = missing_hardwork) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+ggtitle("Non-Responders In The Hardwork Question")+labs(x="Income Levelsk", y="Count")
print(missingh_incomes)
ggsave("income_hardwork_norespondents.jpg")
```
The distribution here is skewed to the left, suggesting that mostly wealthy individuals opted out of the question on hard-work. Hence, my hypothesis was correct. Nepotism is mostly common in high-racial status groups. Hence, it is likely that those neptostic individuals who opted out of the hard-work question were also white. 

```{r, echo=FALSE}
missing_hardwork<- subset(gss,is.na(workhard))
missingh_race<-ggplot(aes(race), data = missing_hardwork) + geom_bar() + ggtitle("Nepotistic Non-Responders in The Hardwork Question")+labs(x="Race of Non-Responders", y="Count")+ theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(missingh_race)
ggsave("race_hardwork_norespondents.jpg")
```

My hypothesis was correct - the overwhelming majority of non-responders to the hard-work question were white individuals. 

###Political Affiliation and Feelings of Proximity to Whites and Blacks 

I am interested in examining the relationship between perceptions of blacks and perceptions of whites. I plotted a scatterplot with a linear smoothing line to observe potentially interesting trends. I hypothesized that this relationship should be most linear, with a slight bend towards proximity to whites as the number of white respondents in the survey were greater than black respondents. 

```{r, echo=FALSE}
set.seed(1)
bwprox<-ggplot(aes(closeblk, closewht), data = gss) + geom_jitter() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Proximity between blacks and whites")+labs(x="Proximity to Blacks", y="Proximity to Whites") + geom_smooth(method="glm")
print(bwprox)
ggsave("bwprox.jpg")
```

The relationship is linear, but it is not a very strong one. As observable, the relationship becomes more linear for individuals who give both blacks and white a higher than 4.5 proximity rating. Therefore, I next examined the relationship between proximity to blacks and whites for those individuals who rated both these proximities as being greater than 5. I hypothesized that in this analysis, the correlation would increase and a more linear relationship would emerge. 

```{r, echo=FALSE}
set.seed(1)
high_proximity_race<-subset(gss, closeblk>=5 & closewht>=5)
high_bwprox<-ggplot(aes(closeblk, closewht), data = high_proximity_race) + geom_jitter() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Proximity between blacks and whites")+labs(x="Proximity to Blacks", y="Proximity to Whites") + geom_smooth(method="glm")
print(high_bwprox)
ggsave("high_bwprox.jpg")
```
My hypothesis was correct - analyzing individuals who rate high proximity to both blacks and whites (greater than 5) reveals that for these individuals, proximities to both races is correlated in a near-linear trend. 


I was also interested in examining how political views related to feelings of proximity with whites. I plot the mean values of white-proximity across individuals of various political views. I hypothesize that as ideologies get more conservative, the perceived proximity to whites increases. 

```{r, echo=FALSE}
averaceprox<-ggplot(aes(polviews, closewht), data = gss) + stat_summary(fun.y = "mean", geom="point") + theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Average Perceived Proximity to Whites Across Political Views")+labs(x="Political Views", y="Average Proximity to Whites")
print(averaceprox)
ggsave("averaceprox.jpg")
```

My hypotehsis was correct: as ideologies span from liberal to conservative, the average reported proximity to whites follows a linear relationship. Next, I examined the data for potential outliers giving the suspiciously high white-proximity value of Extreme Conservatives. 

```{r, echo=FALSE}
pol_raceprox<-ggplot(aes(polviews, closewht), data = gss) + geom_boxplot() + ggtitle("Outlier Detection")+labs(x="Political Views", y="Average Proximity to Whites")+theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(pol_raceprox)
ggsave("pol_raceprox.jpg")
```

I found that there were no real outliers in the data. There is a lot of downward variability in the data, but overall, we don't see observations very far from the mean. 


###Confidence in Science

Lastly, I was interested in examining relations between other variables and the confidence in science variable. First, the general patterns of responses: 

```{r, echo=FALSE}
con_sci<- ggplot(aes(consci), data = gss) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+ggtitle("Confidence in Science")+labs(x="Level of Confidence in Scientific Community", y="Count")
print(con_sci)
ggsave("con_sci.jpg")
```

I was intrigued by the high level of non-responders to the confidence in scientific community question so I examined the demographics of non-responders: 

```{r, echo=FALSE}
science_spend_missing<- subset(gss, is.na(natsci))
science_spend_NA<-ggplot(aes(income06, polviews, color=race), data=gss)+geom_jitter()+theme(axis.text.x = element_text(angle = 90, hjust = 1))+ggtitle("Non-Responders to Confidence in Science")+labs(x="Income", y="Political Views")
print(science_spend_NA)
ggsave("science_spend_NA.jpg")
```
Pretty robust trends in attriton: White moderates from a wide range of income-levels avoided answering the question but there was little to no attrition in political views distant from those of the moderates. 

I finally examined how education levels related to confidence in the scientific community. I hypothesized that people with higher educations levels would typically put greater confidence in the scientific community. 

```{r, echo=FALSE}
science_education<-ggplot(aes(educ, consci), data = gss) + geom_jitter() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Confidence in Science and Education Level")+labs(x="Education Levels", y="Confidence in Science")
print(science_education)
ggsave("science_education.jpg")
```

It looks like the greatest confidence in science is exhibited by individuals who have attended either a community or a full-time college. PhD students and PhD holders do not appear to have a great confidence in the scientific community ironically, perhaps because they are over-integrated with the system and can see the finer faults and vices. 



