---
title: "Assignment4_Simulate_income"
author: "Chen Anhua"
date: "November 3, 2017"
output: md_document
---

```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
rm(list = ls())
options(warn=-1)

```

## Simulating Your Income

### Model set up
We will firstly set up our model to simulate the income.

```{r model, include=TRUE, echo=TRUE}
# ===================
# inc_sim is a function to simulate the income trajectory of all simulations
# ===================

inc_sim = function(inc0, rho, g, sigma, start_time, end_time, ntrials, seed){
  set.seed(seed)
  # set up the matrix, row represents simulations and column represents time
  duration = end_time - start_time + 1
  inc_mat = as.data.frame(matrix(rep(NA, ntrials * duration), nrow = ntrials))
  rownames(inc_mat) = paste0("sim", seq(1 : ntrials))
  colnames(inc_mat) = seq(start_time, end_time)
  
  # simulate the shocks matrix
  eps_mat = as.data.frame(matrix(rnorm(ntrials * duration, 0, sigma), 
                                 nrow = ntrials, byrow = F))
  rownames(eps_mat) = paste0("sim", seq(1 : ntrials))
  colnames(eps_mat) = seq(start_time, end_time)
  
  #initialize the start year's income
  inc_mat[as.character(start_time)] = log(inc0) + 
    eps_mat[as.character(start_time)]
  
  for (year in (start_time + 1):end_time){
    inc_mat[as.character(year)] = 
      (1 - rho) * (log(inc0) + g * (year - start_time)) +
      rho * inc_mat[as.character(year - 1)] +
      eps_mat[as.character(year)]
  }
  
  return(exp(inc_mat))
  }
```

### Question 1

```{r q1, echo=T}
#initialize parameters
inc0 = 80000
rho = 0.2
g = 0.03
sigma = 0.1
start_time = 2019
end_time = 2058
ntrials = 10000
seed = 1
# simulate income trajectories for question 1
inc_mat1 = inc_sim(inc0, rho, g, sigma, start_time, end_time, ntrials, seed)

# pick out a simulated trajectory and plot it
inc_sim1 = as.data.frame(t(inc_mat1[1,]))
inc_sim1$year = as.numeric(rownames(inc_sim1))
plot1 = ggplot(inc_sim1, aes(year, sim1)) +
  geom_line(color = "blue", size = .8) +
  ylab("Simulated income") +
  xlab("Year") +
  ggtitle("Question1: One instance of simulated income trajectories")
print(plot1)


```

### Question 2

As we can see in the chart below, the distribution of simulated start income is symmetric and bell-shaped. As we calculated below, 1.27% of the class will earn more than 100000 in the first year while 9.47% of the class will earn less than 70000 in the first year according to our simulation.

```{r q2, echo=T, include=T}
inc_start = inc_mat1[as.character(start_time)]
plot2 = ggplot(inc_mat1, aes(x = inc_mat1[as.character(start_time)])) +
  geom_histogram(bins = 50) + 
  ylab("Count of simulations") +
  xlab("Simulated start income") +
  ggtitle("Question 2: distribution of simulated start income")

print(plot2)


# calculate the upper and lower tail of the start income distribution
uppertail =  100 * length(inc_start[inc_start > 100000])/ntrials
lowertail =  100 * length(inc_start[inc_start < 70000])/ntrials
print(paste0("Percentage of class earning more than 100000: ", uppertail))
print(paste0("Percentage of class earning less than 70000: ", lowertail))


```


### Question 3

We will define a function to calculate the year when each simulated student payoff his/her debt.

```{r q3, echo=T, include=T}

payoff_year = function(inc_mat, debt, end_time){
  payment_mat = t(apply(inc_mat * 0.1, 1, cumsum))    # calculate cumulative payments
  payment_mat[payment_mat < debt] = 0    # index year where debt is not paid off
  payment_mat[payment_mat >= debt] = 1    # index year where debt is paid off
  payoff_ind = end_time - as.data.frame(rowSums(payment_mat)) + 1    # calculate pay-off year
  colnames(payoff_ind) = "payoff_year"
  return(as.data.frame(payoff_ind))
}

```

We then plot the distribution of simulated students' payoff year. Only 17.98% of the simulated student can payoff their debts on or before 2028.

```{r q3_plot, echo=T, include=T}
payoff_year1 = payoff_year(inc_mat1, 95000, 2058)

plot3 = ggplot(payoff_year1, aes(x = payoff_year1$payoff_year )) +
  geom_histogram(bins = length(unique(payoff_year1$payoff_year))) + 
  ylab("Count of simulated students") +
  xlab("Year where simulated students pay off the debt")+
  ggtitle("Question 3: Distribution of years when debt is paid off")

print(plot3)

percentage_payoff_2028 = length(payoff_year1[payoff_year1 <= 2028])/ntrials * 100
print(paste0("Percentage of students pay off debt on /before 2028: ",
             percentage_payoff_2028))

```

### Question 4

Please see the chart below, and the percentage of simulated students who can pay off student debt is 69.94%. 


```{r q4, echo=T, include=T}
inc0 = 85000
sigma = 0.15
inc_mat2 = inc_sim(inc0, rho, g, sigma, start_time, end_time, ntrials, seed)
payoff_year2 = payoff_year(inc_mat2, 95000, 2058)
plot4 = ggplot(payoff_year2, aes(x = payoff_year2$payoff_year )) +
  geom_histogram(bins = length(unique(payoff_year2$payoff_year))) + 
  ylab("Count of simulated students") +
  xlab("Year where simulated students pay off the debt")+
  ggtitle("Question 4: Distribution of years when debt is paid off")


print(plot4)

percentage_payoff_2028 = length(payoff_year2[payoff_year2 <= 2028])/ntrials * 100
print(paste0("Percentage of students pay off debt on /before 2028: ",
             percentage_payoff_2028))

```







