---
title: "Kaggle Open Call"
author: "Kristopher Nichols"
date: "November 12, 2017"
output:
markdown_document:
  theme: readable
---
#Question 1: Part II

I found the WSDM - KKBox's Music Recommendation Challenge to be very interesting. It is a competition started by the 11th ACM International Conference on Web Search and Data Mining and utilizes data provided by Asia's leading pop station. Using this provided data it tasks the competitors to create a better music reccomendation system than is currently available with computational music prediction techniques. A key problem with music prediction algorithms as currently constituted is that they often lack historical data to make accurate predictions with, and this is a problem the competition specifically targets. As stated, the central task is for competitors to predict the likelihood of a listener listening to a song "repetitively after the first observable listening event within a time window was triggered." The team with the highest predictive score is awarded $2,500 which will be presented at the conference.

Submission for the competition is relatively simple and stripped down. The competition is open to all over the age of 18 and although public sharing is allowed, private sharing of data is not. All that is needed to submit is to register online via kaggle. Competitors are given training data and are encouraged to use outside data to increase their predictive power. 

#Part III
```{r global_options}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
## Load Libraries
```{r}
library(readr)
library(ggplot2)
library(dplyr)
```

##Load Data
```{r}
video_game_sales_with_ratings <-read_csv("~/GitHub/persp-analysis/students/nichols_kristopher/Data for HW5/Video_Games_Sales_as_at_22_Dec_2016.csv")
View(video_game_sales_with_ratings)
```

##Clean Data
```{r}
video_game_sales_with_ratings <-subset(video_game_sales_with_ratings, !is.na(User_Score))
video_game_sales_with_ratings1 <-subset(video_game_sales_with_ratings, User_Score != "tbd")
```

##Make Graph
```{r}
video_game_sales_with_ratings1 %>%
  ggplot(aes(Critic_Score, User_Score, color = Critic_Score)) +
  geom_point() +
  theme_bw() +
  labs(title = "How User Score Relates to Critic Score", x = "Critic Score", y = "User Score", subtitle = "MetaCritic Data", color = "Score")
```

#Question 2: Improving a Journal Article

In the paper "Co-benefits of addressing climate change can motivate action around the world" by Bain et al. [2016](https://repository.up.ac.za/bitstream/handle/2263/51902/Bain_CoBenefits_2016.pdf?sequence=1) researchers looked to uncover which framing of climate change benefits most effectively convinces individuals to enact in pro-environmental behavior. Using a global sample, researchers found that emphasizing on the benevolent (moral caring) and developmental (societally beneficial) co-benefits of climate change, that is, the potential for all to prosper resulted in greater likelihood of action. 

This study is very interesting, but has one major pitfall: although the sample is global, the sample is still not very high in external validity because it solely utilizes university students. This is an issue that could be controlled for very well with a human computation element. For example, I propose that the research team construct websites for each country that is planned to be a part of the study. On this website the researchers will train participants in articulating their feelings in an analyzable way, identifying their political affiliation, and will be prompted to enter in their feelings on climate change. 

Instructions will include that participants should emphasize the most important element about climate change for them, and why policy tackling the issue is good *and* bad. Each website for each country can be peddled by bots on twitter or Facebook, and the website can be trained to search for key phrases such as "it helps everyone" or "too expensive." This data can then be combined, and mapped cumulatively but also piecewise with many different analyses including random forest analysis.

#Question 3: Alternative Assignment
