---
title: "Simulating Income"
author: "Kristopher Nichols"
date: "November 5, 2017"
output:
md_document:
  theme: readable
---
## Load Libraries
```{r setup, echo =FALSE, message =FALSE, error=FALSE, warning =FALSE}
library(tidyverse)
library(knitr)
library(sca)
library(dplyr)
library(ggplot2)
set.seed(4566)
```

## Question 1: Estabshling function, creating object (matrix) for data to be saved to, and preparing variables for graphing

```{r}

income1 <- function(p, g, SD, Years, Year_Start, simulations, 
                 initial_income){ #This line of code establishes the basic variables involved in the function (Given by the assignment).
  error_log <- t(matrix(rnorm(simulations*Years, sd= SD), #Establish error variable for use later in function (in for loop).
                       nrow= Years))
  inc_ln <- matrix(nrow= simulations, ncol= Years) #Create matrix for data to fill out.
  inc_ln[,1] <- log(initial_income) + error_log[,1] 
  for(i in 2 : Years){
    inc_ln[, i] <- (1 - p)*(log(initial_income) + g*(i - 1)) +
      p*inc_ln[, i - 1] + error_log[, i]
  } #Establish value of log income.
    Graph_Year <- c(Year_Start : (Year_Start + Years - 1)) %>% #Reconstitute year variable as Graph_Year to satisfy t-1
    as_data_frame() %>%
    rename(Graph_Year= value)
  Simulated_Income <- inc_ln %>% #Consolidate the established error variable, year, variable, and income variable into Simulated income object
    t() %>%
    as.tibble() %>%
    bind_cols(Graph_Year) %>%
    gather(ID, inc_ln, 1 : simulations) %>%
    mutate(income= exp(inc_ln))
  return(Simulated_Income)
}
#After attempting to use strictly for loops as the homework implied was possibly I was unable to get it to work and ultimately just attemted to use function and specify each variable - this was my result. 
```


##Utilizing specified values in Q1, plugging into equation, and graphing.
```{r Q1, warning =FALSE}

Simulated_Income <- income1(p= 0.2, g= 0.03, SD= 0.1, Years= 40, Year_Start= 2019, simulations= 10000, 
                 initial_income= 80000)

Simulated_Income %>% 
  filter(ID== "V4566") %>% #Use established seed as random person
  ggplot(aes(Graph_Year, income)) +
  geom_line() +
  geom_point(color = "red") +
  labs(x= "Year", y= "Income", title= "Simulated Income from 2019 to 2058", subtitle= "For Randomly Generated Seed") + 
  theme_classic()
```


## Question 2: Filter out year 2019
```{r, warning= FALSE}

Simulated_Income_2019 <- Simulated_Income %>%
  filter(Graph_Year== 2019)
```

##Question 2: Calculate what percent will earn less than $70,000
```{r, warning= FALSE}

income_70 <- percent(nrow(filter(Simulated_Income_2019, income < 70000)) / nrow(Simulated_Income_2019), d= 2)
income_70 %>%
  kable()
```
Will earn an income of less than = 70,000


##Question 2: Calculate what percent will earn more than $100,000
```{r, warning= FALSE}

income_100 <- percent(nrow(filter(Simulated_Income_2019, income > 100000)) / nrow(Simulated_Income_2019), d= 2) 
income_100 %>%
  kable()
```
Will earn an income of more than = $100,000


##Question 2: Graph
```{r, warning =FALSE}

ggplot(Simulated_Income_2019, aes(income)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  stat_function(fun= dnorm, args= list(mean = mean(Simulated_Income_2019$income), sd= sd(Simulated_Income_2019$income))) +
  labs(title= "Distribution of Income Amongst MACSS Students", x= "Income", y = "Density", subtitle= "In 2019") +
  theme_minimal() +
  scale_y_continuous(label= scales::percent) +
  scale_x_continuous()
```




###Graph Explanation:
The distribution maps onto a normal distrbution very easily, matching lmost indetically with a standard bell-curve.



##Question 3: Establish codes for calculating debt 
```{r, warning= FALSE}

Simulated_Debt <- Simulated_Income %>%
  group_by(ID) %>%
  mutate(Simulated_Debt= cumsum(0.1 * income)) %>%
  filter(Simulated_Debt>= 95000) %>%
  group_by(ID) %>%
  filter(Simulated_Debt== min(Simulated_Debt))
```

##Question 3: Calculating percent of people who will pay off loans by 2028 by filtering out 2029+ data.
```{r, warning= FALSE}
debt_success <- percent(nrow(filter(Simulated_Debt, Graph_Year < 2029)) / nrow(Simulated_Debt), d= 2)
debt_success %>%
  kable()
```
Will pay their loans by 2028.

##Question 3: Graph
```{r}

# Graphing
ggplot(Simulated_Debt, aes(Graph_Year, ..density..)) +
  geom_histogram(bins= 3) +
  scale_y_continuous(label= scales::percent) +
  labs(title= "Distribution of When People Pay Their Loans", x= "Year", y= "Density") +
  theme_minimal()
```


###Graph Explanation: 
The histogram illustrates that while most students do not pay off their loans by 2028, the vast majority are able to pay off their loans by 2029 (with a density value approaching 80%)



##Question 4: Create new code that raises initial income to $85,000
```{r, warning= FALSE}

Debt_New_MACSS <- income1(p = 0.2, g = 0.03, SD = 0.15, Years = 40, Year_Start = 2019, simulations = 10000, 
                 initial_income = 85000)

Debt_95 <- Debt_New_MACSS %>%
  group_by(ID) %>%
  mutate(Debt_95 = cumsum(0.1*income)) %>%
  filter(Debt_95 >= 95000) %>%
  group_by(ID) %>%
  filter(Debt_95 == min(Debt_95))

Successful_Debt <- percent(nrow(filter(Debt_95, Graph_Year < 2029)) / nrow(Debt_95), d = 2)
Successful_Debt %>%
  kable()
```
Will pay their loans by 2029.

##Question 4: Graph
```{r, warning= FALSE}
ggplot(Debt_95, aes(Graph_Year, ..density..)) +
  geom_histogram(bins = 5) +
  labs(title = "A Distribution of When MACSS Students Pay off their Loans", x = "Year", y = "Density") +
  theme_minimal() +
  scale_y_continuous(label = scales::percent)
```



###Graph Explanation: 
The histogram with the adjusted initial income illustrates that by 2028 the majority of students are able to pay off their debt by 2028. 