---
title: 'Assignment #4 Simulate Your Income'
author: "Xi Chen"
date: "November 6, 2017"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
## 1. Plot one of the lifetime income paths
```{r}
  set.seed(123)

  # Create a function of stimulating income 
  income_stimulate = function(error_mean, error_sd, rho, g, inc0, years,n_sample){

    #Create an empty matrix to store the stimulated income
    inc_ln = matrix(nrow=years, ncol=n_sample)

    # Create a matrix for log_error terms
    error_ln = matrix( rnorm(n=years*n_sample, mean=error_mean, sd=error_sd),
                       nrow=years, ncol=n_sample)
  
    # The log_income for the year of 2019
    inc_ln[1,] = log(inc0) + error_ln[1,]
  
    # The log_income for the year of 2020 to 2058
    for (t in 2:years){
      inc_ln[t,] = (1-rho)*( log(inc0)+g*(t-1) ) + rho*inc_ln[t-1,] + error_ln[t,]
    }
    inc = exp(inc_ln)
  }

  # Plug prameteres into the function
  inc = income_stimulate(error_mean=0, error_sd=0.1, rho=0.2, g=0.03, 
                         inc0=80000, years=40, n_sample=10000)

  # Plot one of the lifetime income paths
  options(warn=-1)
  library(ggplot2)

  inc = as.data.frame(inc)
  rownames(inc) = c(2019:2058)
  ggplot(inc, aes(rownames(inc), inc[,1], group=1)) + 
    geom_line(color = "red", size=1) +
    geom_point(size = 2) +
    scale_x_discrete(breaks=seq(2019, 2058, 3)) +
    labs(title="The Simulated Annual Income Path (2019 - 2058)",
         x="Year", y="Annual Income ($)") 

```


## 2. Plot a histogram of 2019 income
```{r}
  # Plot a historgram with 50 bins of 2019 Income for 10000 simulations
  inc = as.matrix(inc)
  inc_2019 = inc[1,]
  hist(inc_2019, main="Histogram of Stimulated Income in 2019", 
       xlab="Stimulated Income",breaks=50)
  grid(col="gray")

```
Comments: 
The distribution is very close to a normally distribution. 
However, the right tail is a little bit longer than the left tail, which might indicate right skewness. 

```{r}
  sum(inc_2019>100000)/10000

```
Comments: The percentage of the income which will be more than $100,000 is about 1.24%.

```{r}
  sum(inc_2019<70000)/10000

```
Comments: The percentage of the income which will be less than $70,000 is about 8.94%.

## 3. Plot the histogram of years needed to pay off the debt
```{r}
  total_debt=95000
  pay_rate=0.1
  n_sample=10000
  debt_year = matrix(nrow=n_sample, ncol=1)

  # Compute the years needed to pay off the debt for each simulation
  for (n in 1:n_sample){
  
    stimulate_n = inc[,n]
    payment = 0
    pay_year = 0
  
    while(payment <= total_debt){
      payment = payment + stimulate_n[pay_year+1]*pay_rate
      pay_year = pay_year + 1
    }
    
    debt_year[n, 1] = pay_year
  }
  
  # Histogram of the years to pay off the loan
  debt_year=as.data.frame(debt_year)
  ggplot(debt_year, aes(debt_year)) + 
    geom_histogram() + 
    scale_x_continuous(breaks=seq(9, 13, 1)) + 
    labs(title="Histogram of Years Needed to Pay off the Loan", 
         x="Years", y="Frequency") + stat_bin(bins=5)

```

```{r}
  sum(debt_year<=10)/10000

```
Comments: The percentage that paying off the loan in 10 years in all stimulations is about 18%.

## 4. New stimulations with new parameters
```{r}
  # Plug new parameters into the function of income stimulation
  set.seed(123)
  new_inc = income_stimulate(error_mean=0, error_sd=0.15, rho=0.2, g=0.03,
                             inc0=85000, years=40, n_sample=10000)
  
  # The years it takes to pay off the debt
  new_debt_year = matrix(nrow=n_sample, ncol=1)
  for (n in 1:n_sample){
  
    stimulate_n = new_inc[,n]
    payment = 0
    pay_year = 0
  
    while(payment <= total_debt){
      payment = payment + stimulate_n[pay_year+1]*pay_rate
      pay_year = pay_year + 1
    }
    
    new_debt_year[n, 1] = pay_year
  }

  # Histogram of the years to pay off the loan
  options(warn=-1)
  new_debt_year=as.data.frame(new_debt_year)
  ggplot(new_debt_year, aes(new_debt_year)) + 
    geom_histogram() + 
    scale_x_continuous(breaks=seq(8, 12, 1)) + 
    labs(title="Histogram of Years Needed to Pay off the Loan (New)", 
         x="Years", y="Frequency") + stat_bin(bins=5)

```

```{r}
  sum(new_debt_year<=10)/10000

```
Comments: With the new parameters, the percentage that paying off the loan in 10 years in all simulations is about 70.35%. 

