---
title: "Exploration Write-Up"
author: "Nora Nickels"
date: "11/27/2017"
output: github_document
---

# Assignment 7 - EDA
## Perspectives of Computational Analysis - Fall 2017

# Exploration Write-Up

## Introduction

For my exploration write-up, I focus on two different topics based on what I learned from my exploratory data analysis: 1) how does family size, in terms of children and number of siblings, vary with both religious denomination and reported strength; and 2) what subjective confidence in national systems (e.g., business, medicine) varies with reported subjective happiness?  

```{r, echo=FALSE, message=FALSE}

# Install pacakges and import data. 
# install.packages("poliscidata")
data(gss, package = "poliscidata")

# Convert data to tibble
library(tidyverse)
gss <- as_tibble(gss)

library(tidyverse)
library(knitr)
library(broom)
library(stringr)
library(forcats)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())

```

## Family size and Religion/Faith

```{r Write-Up 1, echo=FALSE}

relig_summary <- gss %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    sibs = mean(sibs, na.rm = TRUE),
    childs = mean(childs, na.rm = TRUE),
    n = n()
  )

relig_summary %>%
  ggplot(aes(fct_reorder(relig, sibs), sibs)) +
    geom_col() +
    coord_flip() +
    labs(title = "Average Family Size by Reported Religion - Siblings",
        y = "Average Number of Siblings",
        x = "Religion")

relig_summary %>%
  ggplot(aes(fct_reorder(relig, childs), childs)) +
    geom_col() +
    coord_flip() +
    labs(title = "Average Family Size by Reported Religion - Children",
        y = "Average number of Children",
        x = "Religion")

```

## What varies with subjective happiness?

```{r Write-Up 2, echo=FALSE}

gss %>%
  count(conbus, happy) %>%
  na.omit() %>%
  ggplot(aes(conbus, n, fill = happy)) +
  geom_col(position = "fill") +
  labs(title = "Subjective Happiness Make-Up of Subjective Confidence in Major Companies",
        x = "Confidence in Major Companies",
        y = "Proportion of Count",
        color = "Subjective Happiness")

```


