---
title: "Simulation_my_income"
author: "Zhiyu Fu"
date: "11/1/2017"
output: md_document
---

# Simulation of My Lifetime Income

```{r, message=TRUE, warning=TRUE, include=FALSE}
library(stringi)
library(data.table)
library(ggplot2)
library(dplyr)
```

## Q1
First we generate a matrix of simulated income. 

```{r}
set.seed(1)
sigma = 0.1
rho = 0.2
g = 0.03
log_inc0 = log(80000)
N = 10000
Y = 40
simulated_income = matrix(0, N, Y)
log_epsilon = rnorm(N, 0, sigma)
simulated_income[,1] = log_inc0 + log_epsilon
for (i in 2:40)
{
  
  log_epsilon = rnorm(N, 0, sigma)
  simulated_income[,i] =  (1-rho)*(log_inc0+g*(i-1)) + rho * simulated_income[,i-1]+ log_epsilon
}
simulated_income = exp(simulated_income)
one_path = simulated_income[sample.int(N, 1),]
qplot(2019:2058, one_path, geom = "line", xlab = "Year", ylab = "Simulated Income", main = "An instance of simulated income")

```
## Q2

1.27% of class will earn more than 100,000 in 2019, while 9.47% of students will earn less than 70,000. 

Theoretically, the simulated income is lognormally distributed. Though the histogram of income is very close to a normal distribution, it is still right-skewed, i.e., it has a long tail on the right. A QQ plot can support that it is not a normal distribution.

```{r}
# ggplot(simulated_income[,1]) + geom_histogram(bins = 200)
qplot(simulated_income[,1], bins = 50, ylab = "Count", xlab = "Simulated Income of year 2019", main = "Distribution of simulated income of year 2019")
print("Percentage of students who earn more than 100000 in 2019")
print(sum(simulated_income[,1]>100000)/N*100)
print("Percentage of students who earn less than 70000 in 2019")
print(sum(simulated_income[,1]<70000)/N*100)
qqnorm(simulated_income[,1])
qqline(simulated_income[,1])
```
## Q3

The histogram is as follows. Only 17.98% students can pay off the loan on or before 2028.


```{r}
cum_sum = t(apply(simulated_income*0.1, 1, cumsum))
pay_off_year = sapply(apply(cum_sum>=95000, 1, which), min) + 2018
qplot(pay_off_year, bins = length(unique(pay_off_year)), 
      xlab = "Years when students pay off their debt",
      ylab = "Count", main = "Distribution of years when students pay off their debt")
print("The percentage of students who can pay off their debt on or before 2028.")
print(sum(pay_off_year<=2028)/N*100)
```
## Q4

With the new parameters, 69.94% of students can pay off their debt on or before 2028.

```{r}
log_inc0 = log(85000)
sigma = 0.15
set.seed(1)
log_epsilon = rnorm(N, 0, sigma)
simulated_income[,1] = log_inc0 + log_epsilon
for (i in 2:40)
{
  
  log_epsilon = rnorm(N, 0, sigma)
  simulated_income[,i] =  (1-rho)*(log_inc0+g*(i-1)) + rho * simulated_income[,i-1]+ log_epsilon
}
simulated_income = exp(simulated_income)
cum_sum = t(apply(simulated_income*0.1, 1, cumsum))
pay_off_year = sapply(apply(cum_sum>=95000, 1, which), min) + 2018
qplot(pay_off_year, bins = length(unique(pay_off_year)), 
      xlab = "Years when students pay off their debt",
      ylab = "Count", main = "Distribution of years when students pay off their debt")
print("The percentage of students who can pay off their debt on or before 2028.")
print(sum(pay_off_year<=2028)/N*100)
```

