---
title: "assignment_8"
output: github_document
---

###Colleges 

#### Question 1 

Perform PCA analysis on the college dataset and plot the first two principal components. Describe the results.

```{r chunk 1, include=FALSE}
library("dplyr")
library("ggfortify")
library("dendextend")
data<-read.csv("College.csv")
data.numeric<-select_if(data, is.numeric)
data.pca<-prcomp(data.numeric, center = TRUE, scale.=TRUE)
plot<-autoplot(data.pca, loadings=TRUE, loadings.label=TRUE)
```

Describe the results:

####Question 2 

```{r colleges-Q2, include=FALSE}
summary(data.pca)
```

##USA Arrests

###Question 1

```{r USA-1, include=FALSE}
data_usa<-read.csv("USArrests.csv")
data_usa.numeric<-select_if(data_usa, is.numeric)
pca.data_usa<-prcomp(data_usa.numeric, center=TRUE, scale.=TRUE)
plot_usa<-autoplot(pca.data_usa, loadings=TRUE, loadings.label=TRUE)
```
Describe your results.

###Question 2

```{r USA-2, include=FALSE}
plot_kmeans<-autoplot(kmeans(USArrests, 2), data = USArrests)
```
Describe your results.

###Question 3 
```{r USA-3, include=FALSE}
plot_kmeans_Q3<-autoplot(kmeans(USArrests, 4), data = USArrests)
```

###Question 4
```{r USA-4, include=FALSE}
plot_kmeans_Q4<-autoplot(kmeans(USArrests, 3), data = USArrests)
```

###Question 5
```{r USA-5, include=FALSE}
pca_scores<-data.frame(pca.data_usa$x)
pca_scores_plot<-data.frame(subset(pca_scores, select=c("PC1", "PC2")))
plot_kmeans_Q5<-autoplot(kmeans(pca_scores_plot, 3), data =pca_scores_plot)
```
Describe your results and compare them to the clustering results with $K=3$ based on the raw data.

###Question 6 
```{r USA-6, include=FALSE}
cluster<-hclust(dist(USArrests))
```

###Question 7
```{r USA-7, include=FALSE}
chop_cluster<-data.frame(cutree(cluster, 3))
```

Which states belong to which clusters?
```{r chop print}
print(chop_cluster)
```

###Question 8
```{r USA-8, include=FALSE}
chop_scaled<-scale(USArrests)
cluster_scaled<-hclust(dist(chop_scaled))
cluster_scaled_chop<-data.frame(cutree(cluster_scaled,3))
print(cluster_scaled_chop)
```
What effect does scaling the variables have on the hierarchical clustering obtained? 
