---
title: "yu_liqiang_EDA_Lab_Notebook"
author: "Liqiang Yu"
date: "11/26/2017"
output: github_document
---
###Load Data
```{r echo=FALSE}
library(poliscidata)
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(GGally)
data(gss, package = "poliscidata")
gss <- as_tibble(gss)
```

###Gathering basic information about our data for later use
```{r results='hide'}
summary(gss) ##Result hidden
```

#Variable variation exploration

###Race distribution for this survey
```{r}
gss %>%

  ggplot(aes(race))+
  geom_bar()
```

### Sex distribution for this survey
```{r}
gss %>%

  ggplot(aes(sex))+
  geom_bar()
```

### Age distribution for interviewee
```{r}
summary(gss$age)
gss %>%

  ggplot(aes(age))+
  geom_density()

```

###Age dis among sex groups
```{r}
ggplot(gss, aes(age, color = sex)) +
  geom_density()
```

### Work status 
```{r}

gss %>%

  ggplot(aes(wrkstat))+
  geom_bar()
```

### Education status 
```{r}

gss %>%

  ggplot(aes(educ))+
  geom_bar()
```

### People's opinion about gunlaw (require a person to obtain a police permit before he or she could buy a gun)?
```{r}
gss %>%

  ggplot(aes(gunlaw))+
  geom_bar()
```

###How would you say things are these days, very happy, pretty happy, or not too happy?
```{r}
gss %>%

  ggplot(aes(happy))+
  geom_bar()
```

###Violin plot of age
```{r}
ggplot(gss, aes("", age)) +
  geom_violin() +
  labs(title = "Violin plot of age for gss data",
       x = NULL,
       y = "Age")
```

#Outliers

### Extract variables we are interested in
```{r}
gssr <- gss[, c("age", "sex", "marital", "educ", "income06", "happy")]
gssr <- within(gssr,{
    age1 <- as.numeric(age) 
    educ1 <- as.numeric(educ) 
    sex1 <- as.numeric(sex)  
    #code income into categories
    cincome <- as.numeric(income06) })
ggpairs(gssr[,c("age1", "educ1", "sex1", "cincome")],diag = list(continuous = "bar"),lower = list(continuous = "smooth"))
```

### Check outliers for age
```{r}

ggplot(gss,aes("",age)) +
  geom_boxplot() 
summary(gss$age)
```

### Outliers for education
```{r}
ggplot(gssr,aes("",educ1)) +
  geom_boxplot() 
summary(gssr$educ1) #Further investigation for outliers
ggplot(gssr,aes(educ1)) +
  geom_histogram(binwidth  = 1) #Outliers most in <7.5 range
```

### Outliers for income
```{r}
ggplot(gssr,aes("",cincome)) +
  geom_boxplot() 

```

#Covariation


###Heatmap
```{r}
cormat_heatmap <- function(data){
  # generate correlation matrix
  cormat <- round(cor(data), 2)
  
  # melt into a tidy table
  get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
  }
  
  upper_tri <- get_upper_tri(cormat)
  
  # reorder matrix based on coefficient value
  reorder_cormat <- function(cormat){
    # Use correlation between variables as distance
    dd <- as.dist((1-cormat)/2)
    hc <- hclust(dd)
    cormat <-cormat[hc$order, hc$order]
  }
  
  cormat <- reorder_cormat(cormat)
  upper_tri <- get_upper_tri(cormat)
  
  # Melt the correlation matrix
  melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)
  
  # Create a ggheatmap
  ggheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
    geom_tile(color = "white")+
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         midpoint = 0, limit = c(-1,1), space = "Lab", 
                         name="Pearson\nCorrelation") +
    theme_minimal()+ # minimal theme
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                     size = 12, hjust = 1))+
    coord_fixed()
  
  # add correlation values to graph
  ggheatmap + 
    geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid.major = element_blank(),
      panel.background = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "bottom")
}
gssr_tidy <- gssr %>%
  na.omit()
  cormat_heatmap(select_if(gssr_tidy, is.numeric))
```

###Density estimation of income
```{r}
summary(gssr$cincome)
gssr %>%

  ggplot(aes(cincome))+
  geom_density(aes(color = "Gaussian"), kernel = "gaussian") +
  geom_density(aes(color = "Epanechnikov"), kernel = "epanechnikov") +
  geom_density(aes(color = "Rectangular"), kernel = "rectangular") +
  geom_density(aes(color = "Triangular"), kernel = "triangular") +
  geom_density(aes(color = "Biweight"), kernel = "biweight") +
  labs(title = "Density estimator of income level for gss data",
       x = "Income level",
       y = "Density")
```

###Income and education relation
```{r}
ggplot(gssr,aes(educ1,cincome))+ geom_col()+coord_flip()
```

### Jitter and smooth line between income and education
```{r}
ggplot(gssr,aes(educ1,cincome))+ geom_jitter()+geom_smooth()

```

### Jitter and smooth line between income and education (more clearly)
```{r}
gssr <- within(gssr, {reduc <- pmin(pmax(educ1, 9), 18) })
ggplot(gssr,aes(reduc,cincome))+ geom_jitter()+geom_smooth()
ggplot(gssr,aes(reduc,cincome))+ geom_jitter()+geom_smooth()+facet_wrap(~ sex, scales = "free_y")
```

### Jitter and smooth line between income and age
```{r}
ggplot(gssr,aes(age1,cincome))+ geom_jitter()+geom_smooth()
```

###Education distribution between sex group
```{r}
gssr_median <- gssr %>%
  group_by(sex1) %>%
  summarize(median = median(educ1,na.rm = T))
ggplot(gssr) +
  geom_histogram(aes(educ1), binwidth = 1) +
  geom_vline(data = gssr_median, aes(xintercept = median), color = "red") +
  facet_wrap(~ sex1, ncol = 1)
```

###Income distribution between sex group
```{r}
gssr_median2 <- gssr %>%
  group_by(sex1) %>%
  summarize(median = median(cincome,na.rm = T))
ggplot(gssr) +
  geom_histogram(aes(cincome), binwidth = 1) +
  geom_vline(data = gssr_median2, aes(xintercept = median), color = "red") +
  facet_wrap(~ sex1, ncol = 1)
```

###Happiness among different sex groups
```{r}
gss %>%
  count(happy, sex) %>%
  na.omit() %>%
  mutate(happy = factor(happy, levels = c("VERY HAPPY", "PRETTY HAPPY",
                                            "NOT TOO HAPPY", "NA"))) %>%
  ggplot(aes(happy, n, fill = sex)) +
  geom_col()

```

###Happiness among different work status
```{r}
gss %>%
  count(happy, wrkstat) %>%
  na.omit() %>%
  mutate(happy = factor(happy, levels = c("VERY HAPPY", "PRETTY HAPPY",
                                            "NOT TOO HAPPY", "NA"))) %>%
  ggplot(aes(happy, n, fill = wrkstat)) +
  geom_col()
gss %>%
  count(happy, wrkstat) %>%
  na.omit() %>%
  mutate(happy = factor(happy, levels = c("VERY HAPPY", "PRETTY HAPPY",
                                            "NOT TOO HAPPY", "NA"))) %>%
  ggplot(aes(happy, n, fill = wrkstat)) +
  geom_col(position = "dodge")
```

###Work status among different sex groups
```{r}
gss %>%
  count(wrkstat, sex) %>%
  na.omit() %>%
  mutate(wrkstat = factor(wrkstat, levels = c("WORKING FULL TIME","RETIRED","WORKING PART TIME", "UNEMPLOYED, LAID OFF" ,"TEMP NOT WORKING" ))) %>%
  ggplot(aes(wrkstat, n, fill = sex)) +
  geom_col()
```

### Income level among different age groups and work status groups (219 missing values)
```{r}
ggplot(data = gss, mapping = aes(x = age, y = as.numeric(income06), color = wrkstat)) +
  geom_point()
```






