---
title: "Colleges"
author: "Nora Nickels"
date: "12/4/2017"
output: github_document
---

# Assignment 8 - Unsupervised data
## Perspectives of Computational Analysis - Fall 2017

# Colleges

```{r colleges, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install packages
#install.packages("ggfortify")
#install.packages("ISLR")
#install.packages("ggdendro")

# Load packages
library(tidyverse)
library(knitr)
library(forcats)
library(broom)
library(modelr)
library(stringr)
library(ISLR)
library(titanic)
library(rcfss)
library(grid)
library(gridExtra)
library(ggdendro)
library(tidytext)
library(tm)
library(topicmodels)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())

# Load data
Colleges <- read.csv("College.csv")

# Get rid of private variable (non numeric)
Colleges$Private <- NULL

```

### 1. Perform PCA analysis on the college dataset and plot the first two principal components. Describe the results.

**What variables appear strongly correlated on the first principal component?**

**What about the second principal component?**

```{r colleges_PCA_1}

# Perform PCA on Colleges data set.
CollegesPCA <- prcomp(Colleges,
                 scale = TRUE) 

head(CollegesPCA$x)

# Plot the first two principle components.
CollegesPCA$rotation
biplot(CollegesPCA, scale = 0, cex = .6)

```

The first principal component seems to be most strongly correlated with 'PhD', 'Terminal', 'Top10perc', 'Top25perc', and 'Expend'. The second principal component seems to be most strongly correlated, negatively, with 'Apps', 'Accept', 'P.Undergrad', 'F.undergrad', and 'Enroll', and fairly strongly positive correlated with 'perc.Alumni' and 'Outstate'.

### 2. Calculate the cumulative proportion of variance explained by all the principal components (see 10.2.3 in ISLR). 

```{r Colleges_PCA_2}

# Summarizing importance of PCs.
summary(CollegesPCA)

# Plot of variances associated with PCs.
plot(CollegesPCA, type = "l")

```

**Approximately how much of the variance in College is explained by the first two principal components?**

Approximately 58.3% of the variance is explained by the first two principal components. This is a fairly large amount of importance, especially when compared to the components beyond the first two, as seen by the variances graph.


