---
title: "eda"
author: "Jo Denby"
date: '`r Sys.Date()`'
output: pdf_document

---
```{r, message=FALSE}
library(tidyverse)
library(poliscidata)
library(forcats)
```

```{r}
data(gss, package = 'poliscidata')
gss <- as_tibble(gss)
```

I'm first interested in the distribution of education. 
```{r}
ggplot(gss, aes(x = gss$educ)) +
  geom_bar()

ggplot(gss, aes(x=gss$degree)) + 
  geom_bar()
```
There are three peaks - a large portion of people completed high school (and no more), 2 years of college (i.e,. an associates degree), and 4 years of college (i.e., a bachelor's degree).  

Does degree correlate with income in any way?
```{r}
ggplot(data = gss, aes(x=gss$degree, y = gss$income06)) + 
  geom_jitter()

ggplot(data = gss, aes(income06)) +
  geom_bar() + 
  facet_wrap(~ degree)

```
A greater number of respondents have solely HS degrees, so it is hard to compare across degrees. However, it appears that highest degree has a relationship with income.

```{r}
ggplot(data = gss, aes(income06)) +
  geom_bar(aes(y=(..count..)/sum(..count..))) + 
  facet_wrap(~ degree)

```
Let's investigate people with just HS degrees.

```{r}

ggplot(data = gss[gss$degree=='HS',], aes(income06)) +
  geom_bar()

ggplot(data = gss[gss$degree=='HS',], aes(income06)) +
  geom_bar(aes(y=(..count..)/sum(..count..)))

```
These two plots include NA values - let's try to drop them.
```{r}
gss_no_income_na <- gss[!is.na(gss$income06),]

ggplot(data = gss_no_income_na, aes(income06)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(size = 5))

ggplot(data = gss_no_income_na[gss_no_income_na$degree=='HS',], aes(income06)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(size = 5)) + 
  ggtitle('Income Distribution of People with Solely HS Degree')

ggplot(data = gss_no_income_na[gss_no_income_na$degree=='HS',], aes(income06)) + 
  geom_bar(aes(y=(..count..)/sum(..count..))) + 
  theme(axis.text.x = element_text(size = 5)) + 
  ggtitle('Income Distribution of People with Solely HS Degree')

```
The income distribution of HS degree holders is left skewed, with a center around the \$35,000 to \$50,000 range.

Let's compare distributions of HS degree holders and Bachelor's degree holders.

```{r}
hs_bach_gss <- gss_no_income_na[gss_no_income_na$degree=='HS' | gss_no_income_na$degree=='Bachelor deg', ]

ggplot(data = hs_bach_gss, aes(income06)) + 
  geom_bar(aes(y=(..count..)/sum(..count..))) + 
  facet_wrap(~ degree) + 
  theme(axis.text.x = element_text(size = 5))

bach_gss <- gss_no_income_na[gss_no_income_na$degree=='Bachelor deg',]
hs_gss <- gss_no_income_na[gss_no_income_na$degree=='HS',]

###This is a good one!
ggplot() + 
  geom_bar(data = hs_gss, aes(x=income06, y=(..count..)/sum(..count..))) + 
  geom_bar(data = bach_gss, aes(x=income06, y=(..count..)/sum(..count..))) + 
  facet_wrap(~ degree) +
  theme(axis.text.x = element_text(size= 6)) + 
  labs(x="Income ($k)",
       y="Density",
       title = "Income Distribution by Highest Degree") +
  scale_x_discrete(breaks = inc_levels,
                   labels= c('<1','1','3','4','5','6','7','8','10','12.5','15','17.5','20','22.5','25','30','35','40','50','60','75','90','110','130','150+')) 

```
By comparing the two distributions, a few insights are gleaned:
- Both distributions are left-skewed.
- The proportion of Bachelor's degree holders that make >= $150,000 is much greater than that of HS degree holders.
- The Bachelor's degree distribution is centered much higher than the HS degree distribution.


Now let's look at income distribution across ages. This is the age distribution of respondents.
```{r}
ggplot(data = gss_no_income_na,
       aes(age)) + 
  geom_histogram()
```
Need to eliminate the NA values and investigate relationship between age and income.
```{r}
income_age_summary <- gss_no_income_na %>%
  group_by(income06) %>%
  summarise(age = mean(age, na.rm = TRUE))

ggplot(income_age_summary, aes(age, income06)) + 
  geom_point()

ggplot(income_age_summary, aes(age, fct_reorder(income06, age))) + 
  geom_point()

```
The first plot is better - reordering by income makes the plot harder to read. 
For most income levels, the mean age is between 45 and 50 years. 

MODE OF INCOME BY DEGREE


LOOK AT BIBLE BELIEFS BY RACE / DEGREE
