---
title: "IGN Review Data"
author: "Joseph Denby"
---

Kaggle has a dataset for over 18,000 IGN video game reviews [here](https://www.kaggle.com/egrinstein/20-years-of-games). Through this dataset, we can investigate the average game score over time to see how the industry and/or review process have changed.
```{r}
library(tidyverse)
```
Group the data by the game's year of release to calculate the average score per year. I dropped the year 1970 because those games were reviewed specifically as 'classics' so scores from that year are likely positively skewed.
```{r}
ign <- read_csv("ign.csv", col_names = TRUE) 
ign_avg_yr_review <- ign %>%
  group_by(release_year) %>%
  summarise(avg_yr_score = mean(score)) %>%
  filter(release_year != 1970)
```
I fit a smooth curve to the two variables to get a finer look at how average scores have changed over time.
```{r}
ggplot(ign_avg_yr_review, aes(release_year, avg_yr_score)) +
  geom_smooth(method = 'loess') +
  labs(x = 'Release Year', 
       y = 'Average Score (out of 10)', 
       title = 'Average IGN Game Review Scores ',
       caption = "based on data from Kaggle")
```
Based on the plot, it appears that average game scores have generally trended upward since IGN's founding. Except for a portion of the early 2000's, IGN has tended to review more and more positively. 
