---
title: "YuLiqiangHw4"
author: "Liqiang Yu"
date: "11/1/2017"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1


```{r }
library(scales)
sigma = 0.1
mean = 0
inc0 = 80000
rho = 0.2
t = 2058
g = 0.03
inc = list()
inc0_ln = log(inc0)
##Sampling error
error <- matrix(rnorm(400000, mean, sigma),10000,40)
##Randomly choose a set of errors from 10,000 sets to plot 
rdm_num <- sample(1:10000, 1)
##Initial condition
inc <- inc0_ln + error[rdm_num,1]
## Define income function
income.f = function(inc0_ln,inc,g,rho,t,error,rdm_num){

  result=(1-rho)*(inc0_ln+g*(t-2019))+rho*inc[t-2019]+error[rdm_num,t-2018]
  return(result)
}
#Simulate one example
for (i in 2020:2058){
  inc[i-2018] <- income.f(inc0_ln,inc,g,rho,i,error,rdm_num)
}
#Simulate 10,000 example
for (n in 1:10000){
  
  inc1 = rep.int(0,10000)
  inc1 <- inc0_ln + error[n,1]
  for (i in 2020:2058){
  inc1[i-2018] <- income.f(inc0_ln,inc1,g,rho,i,error,n)

  }
  
}
#Plot
inct <- (exp(inc))
year <- c(2019:2058)
plot(year,inct, main = "Lifetime Income Path", xlab="Year", ylab="Income in dollars")
lines(year,inct)
```
Refer to the graph above.

## Q2



```{r echo=FALSE }
inc_0 = rep(0,10000)
for (i in 1:10000){
  inc_0[i] <- inc0_ln+error[i,1]
}


hist(exp(inc_0), breaks=50, main = "Income in 2019", xlab = "Income in dollars")
inc_0_exp <- exp(inc_0)
#Percentage of earning more than $100,000
sum(inc_0_exp>100000)/10000
#Percentage of earning less than $70,000
sum(inc_0_exp<70000)/10000

```
The percent of MACSS class that earn more than \$100,000 is `r percent(sum(inc_0_exp>100000)/10000)`, while less than \$70,000 is `r percent(sum(inc_0_exp<70000)/10000)`. From the graph, we can easily see that the histogram is both bell-curved and symmetric, so it is a nomral distribution.

## Q3

```{r}
debt = 95000
rate_payoff = 0.1
year = rep.int(0,10000)
            
for (n in 1:10000){
  
  inc3 = rep.int(0,10000)
  inc3 <- inc0_ln + error[n,1]
  debt_paying = debt - rate_payoff*exp(inc3[1])
  year[n] = year[n] + 1
  for (i in 2020:2058){
  if (debt_paying <= 0) break
  inc3[i-2018] <- income.f(inc0_ln,inc3,g,rho,i,error,n)
  debt_paying = debt_paying - rate_payoff * exp(inc3[i-2018])
  year[n] = year[n] + 1

  }


  
}
#Percent of students who can pay off in 10 years
  sum(year<=10)/10000
  hist(year, main = "Years needed to pay off loan")
  


```
See the plot above. The percent of students in MACSS class who is able to pay off the loan in 10 years is `r percent(sum(year<=10)/10000)`.

## Q4

```{r}
sigma4 = 0.15
mean4 = 0
inc_new4 = 85000
rho = 0.2
t = 2058
g = 0.03
inc4_ln = log(inc_new4)
##Sampling error with sigma = 0.15
error1 <- matrix(rnorm(400000, mean4, sigma4),10000,40)


debt = 95000
rate_payoff = 0.1
year1 = rep.int(0,10000)
            
for (n in 1:10000){
  
  inc4 = rep.int(0,40)
  inc4 <- inc4_ln + error1[n,1]
  debt_paying = debt - rate_payoff*exp(inc4[1])
  year1[n] = year1[n] + 1
  for (i in 2020:2058){
  if (debt_paying <= 0) break
  inc4[i-2018] <- income.f(inc4_ln,inc4,g,rho,i,error1,n)
  debt_paying = debt_paying - rate_payoff * exp(inc4[i-2018])
  year1[n] = year1[n] + 1

  }

}
#Percentage
  sum(year1<=10)/10000
  hist(year1, main = "Years needed to pay off loan", xlab="Year")
```
From the plot above, the percent is `r percent(sum(year1<=10)/10000)` under new condition.
