---
title: "assignment5"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Kaggle open call projects

#### 2.

I selected an object called "Titanic: Machine Learning from Disaster". It is a dataset about the demographic information of passengers on Titanic, including their survival, ticket class, name, gender, age, the number of siblings or spouses aboard, the number of parents or children aboard, ticket number, passenger fare, cabin number and port of embarkation. There are two csv files in the dataset, one for training, and the other one for testing.

The goal of the competition is to predict which kinds of people are more likely to survive using the method of machine learning. In order to make a submission, the participant has to predict a 0 or 1 value for survival for every passenger in the test dataset.


#### 3.
I will use the "Silicon Valley Diversity Data" dataset for this question. Please open the pdf called "Sillicon_Valley" for the plot.


```{r}


setwd("E:/Chicago/persp-analysis/students/liao_andi/assignment5")
df = read.csv("Reveal_EEO1_for_2016.csv")


# select the useful information
df_use = subset(df, df$count != 0 & df$count != "na" & df$job_category != "Totals" & df$job_category != "Previous_totals" & df$race != "Overall_totals")

df_use$count = as.numeric(df_use$count)


# plot the distribution
library(ggplot2)
attach(df_use)

ggplot(df_use, aes(x = race, y = count, fill = company )) + 
  geom_bar(stat = "identity") +
  labs(title = " The distribution of races for companies in Silicon Valley", x = "race", y = "num of workers") +
   theme(plot.title = element_text(hjust = 0.5)) +
   scale_x_discrete(labels = c("American Indian", "Asian", "Afirican American", "Latino", "Hawaiian", "Mixed Races", "White"))


# generate pdf
pdf("Sillicon_Valley.pdf", width = 10, height = 10)

ggplot(df_use, aes(x = race, y = count, fill = company )) + 
  geom_bar(stat = "identity") +
  labs(title = " The distribution of races for companies in Silicon Valley", x = "race", y = "num of workers") +
   theme(plot.title = element_text(hjust = 0.5)) +
   scale_x_discrete(labels = c("American Indian", "Asian", "Afirican American", "Latino", "Hawaiian", "Mixed Races", "White"))

dev.off()

```

## Improving a journal article

I choose a journal article called "Facial expressions of emotion are not
culturally universal".

In this reseach, researchers recruited 15 Western Caucasian and 15 East Asian observers to categorize 4800 trials of facial stimulation. Their goal is to prove that people from different culture background have different understanding of universal facial expressions of emotions (happy, surprise, fear, disgust, anger, and sad).

In the experiment, each stimulation is a set of facial movement along with mental representation randomly generated by a computer graphics system based on real-person facial models. Observers are required to evaluate two aspects about each stimulation: emotion (happy, surprise, fear, disgust, anger, and sad), and intensity (five-point scale, from ¡°very low¡± to ¡°very high¡±).

Actually, the data collection process can be reformulated as a human computation project. 4800 stimulations for one observer - that's a lot of work. By attracting as much as participants from different culture background to the experiment, not only can we decrease the workload of every participants, but also we are able to increase the diversity of the data.

**So, how can the study be converted into a human computation project?**

After the computer graphics system is established, we should further improve it so that it can be quickly loaded and run in the web-brower environment. The link will be advertised as "Do you know how similar you are to people in your country? Come and join our little game today!" I belive that this interesting title can attact many people to participate in the study even without other rewards.

Anyone knows the link can participate in the experiment without visiting the lab. To start the little website game, the only thing they have to do is to provide their nationality and the region they grow up.

Instead of asking one observer to repeat the experimental procedure for 4800 times, we can redesign each section as presenting 10 stimulatiions. However, the basic task for each trial will stay the same: categorizing the emotion and intensity. It's okay if participants want to exit at any moment, but their data of finished trials will be collected. 

After finishing a section, participants can choose to continue with another section, or leave the interface. No matter what they choose, they will get feedbacks about their performance of last section. The feedback will show that how similar they are to people from different countries.

**So, how can the human computation approach impove the study?**

The human computation project will imporve the study from three aspects.

* Increased sample size/efficiency: Since that there are more people participating in the survey, the statistical inference from the sample will be more powerful and robust. Moreover, it will take less time to collect the same amount of data because more people are contributing to the whole dataset.

* Diversity of sample: Compared with the original study, the sample will be far more diverse. We can't guarantee that people from every country will join our experiment, but we can ensure that people from more than two countries will be attracted to our experiment. Given the increasing diversity of sample, we are able to discover more interesting facts about human perception of emotions in a cross-culture setting.

* More accurancy: It's easy to imagine that 30 observers in the original version of study will get tired at some point, as they have to go through 4800 faces one by one. Their answers might become less and less accurate as time goes by. However, by diving the huge task to a large number of participants, we are confident that everyone can fully focus on the task during that 10 trials. 

To sum up, the data collection will be more efficient and accurate because of the method of human computation project. Therefore, the conclusion will be more convincing.
